### 2024-12-01 parts/appendix.tex
pushing to arxiv

    Identifying the coefficients of the two polynomials (which is possible
    because the products of binomial coefficients translated by $K$ form a
    basis of $\Rel[\vec{X}]$), we conclude that for all $\card{\alpha} \leq d$,
    $n c_{\alpha} = f_{\alpha}$. In particular,
    for all $\card{\alpha} \leq d$, $c_{\alpha} \in \Nat$,
    hence $P$ is a \kl{natural binomial polynomial}.

>>>>

    Identifying the coefficients of the two polynomials (which is possible
    because the products of binomial coefficients translated by $K$ form a
    basis of $\Rat[\vec{X}]$), we conclude that for all $\card{\alpha} \leq d$,
    $n c_{\alpha} = f_{\alpha}$. In particular,
    for all $\card{\alpha} \leq d$, $c_{\alpha} \in \Nat$,
    hence $P$ is a \kl{natural binomial polynomial}.

### 2024-12-01 parts/prelim.tex
pushing to arxiv

\AP A \intro{monomial} is a product of indeterminates and integers. For
instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, $-Y$ is a
\kl{monomial}, but $X + Y$ and $2X^2 + XY$ are not. Every polynomial $P \in
\Rel[X_1, \dots, X_n]$ decomposes uniquely into a sum of \kl{monomials}. A
\kl{monomial} $S$ \intro{divides} a \kl{monomial} $T$, when $S$ divides $T$
seen as polynomials in $\Rat$. For instance, $2X$ \kl{divides} $XY$, $-YZ$
\kl{divides} $X^2 Y Z^3$, and $Y$ does not \kl{divide} $X$. In the
decomposition of $P \in \Rel[X_1, \dots, X_k]$, a \kl{monomial} is a
\intro{maximal monomial} if it is a maximal element for the \kl{divisibility
preordering} of \kl{monomials}. In the polynomial $P(X,Y) \defined X^2 - 2XY + Y^2
+ X + Y$, the set of \reintro{maximal monomials} is $\set{X^2,  -2 XY,  Y^2}$.
In the polynomial $P(X,Y) \defined (X - Y)^2$, the \kl{non-negative}
\kl{monomials} are $X^2$ and $Y^2$.

@@

\begin{example}
    \label{q-polynomial-n-poly:ex}
    Let $f \colon \Sigma^* \to \Nat$ that maps a word $w$
    to the number of distinct pairs of positions in $w$,
    i.e., $f(w) = \binom{\card{w}}{2} = \card{w}(\card{w} - 1) / 2$.
    Then, $f \in \NSF[2]$.
\end{example}
\begin{proof}
    Let us remark that the set $P_w$ of distinct pairs of positions
    $i < j$ in a word $w$ is in bijection with the set $D_w$ of
    decompositons of the form $w = xyz$,
    where $x$ and $y$ are non-empty, 
    via the map $(i,j) \mapsto (w_{1,i}, w_{i+1,j}, w_{j+1,\card{w}})$.
    Let us write $M \defined (\set{0,1}, \max)$
    which is a finite aperiodic monoid, and $\mu \colon \Sigma^* \to M$
    that maps the empty word $\varepsilon$ to $0$ and the other words to $1$.
    Then, let us define $\pi \colon M^3 \to \Nat$
    via $\pi(x,y,z) = x \times y$.
    We conclude because:
    \begin{align*}
        \pi^\dagger(w) 
        &\defined
        \sum_{xyz = w} \pi(\mu(x), \mu(y), \mu(z)) 
        = 
        \sum_{xyz = w \wedge x \neq \varepsilon \wedge y \neq \varepsilon} 1
        = 
        \card{D_w}
        =
        \card{P_w}
        = f(w) \quad .
        \qedhere
    \end{align*}
\end{proof}

>>>>

\AP A \intro{monomial} is a product of indeterminates and integers. For
instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, $-Y$ is a
\kl{monomial}, but $X + Y$ and $2X^2 + XY$ are not. Every polynomial $P \in
\Rel[X_1, \dots, X_n]$ decomposes uniquely into a sum of \kl{monomials}. A
\kl{monomial} $S$ \intro{divides} a \kl{monomial} $T$, when $S$ divides $T$
seen as polynomials in $\Rat$. For instance, $2X$ \kl{divides} $XY$, $-YZ$
\kl{divides} $X^2 Y Z^3$, and $Y$ does not \kl{divide} $X$. In the
decomposition of $P \in \Rel[X_1, \dots, X_k]$, a \kl{monomial} is a
\intro{maximal monomial} if it is a maximal element for the \kl{divisibility
preordering} of \kl{monomials}. In the polynomial $P(X,Y) \defined X^2 - 2XY +
Y^2 + X + Y$, the set of \reintro{maximal monomials} is $\set{X^2,  -2 XY,
Y^2}$. For instance, the \kl{non-negative} \kl{monomials} of $P(X,Y) \defined
(X - Y)^2$ are $X^2$ and $Y^2$.

@@

\begin{example}
    \label{q-polynomial-n-poly:ex}
    Let $f \colon \Sigma^* \to \Nat$ be the function that maps a word $w$
    to the number of distinct pairs of positions in $w$,
    i.e., $f(w) = \binom{\card{w}}{2} = \card{w}(\card{w} - 1) / 2$.
    Then, $f \in \NSF[2]$.
\end{example}
\begin{proof}
    Let us remark that the set $P_w$ of distinct pairs of positions
    $i < j$ in a word $w$ is in bijection with the set $D_w$ of
    decompositions of the form $w = xyz$,
    where $x$ and $y$ are non-empty, 
    via the map $(i,j) \mapsto (w_{1,i}, w_{i+1,j}, w_{j+1,\card{w}})$.
    Let us write $M \defined (\set{0,1}, \max)$
    which is a finite aperiodic monoid, and $\mu \colon \Sigma^* \to M$
    that maps the empty word $\varepsilon$ to $0$ and the other words to $1$.
    Then, let us define $\pi \colon M^3 \to \Nat$
    via $\pi(x,y,z) = x \times y$.
    We conclude because:
    \begin{align*}
        \pi^\dagger(w) 
        &\defined
        \sum_{xyz = w} \pi(\mu(x), \mu(y), \mu(z)) 
        = 
        \sum_{xyz = w \wedge x \neq \varepsilon \wedge y \neq \varepsilon} 1
        = 
        \card{D_w}
        =
        \card{P_w}
        = f(w) \quad .
        \qedhere
    \end{align*}
\end{proof}

### 2024-12-01 parts/starfree.tex
pushing to arxiv

    Note that the same proof scheme can be used to conclude that
    \cref{f-npoly-combi:item} implies
    \cref{f-combinatorial:item}. For the converse implication, we are
    going to introduce $\omega_2$ associated to the fact that $f$ is
    \kl{$(k,\Nat)$-combinatorial}. Because polynomials \kl{represented} by
    \kl{$\Nat$-polyregular functions} and \kl{integer binomial polynomials} are
    both closed under multiplication by a constant factor, we can assume that
    $\omega = \omega_2$ in the decomposition of $f$.
    Now, consider $(S, \vec{r}) \in \ModuloTypes[\omega]^k$. 
    Notice that for all vectors $\vec{x} \in (\Nat_{\geq 1})^k$, 
    the vector $(x_1 \omega \ind{S}(1) + r_1, \dots, x_k \omega \ind{S}(k) + r_k)$ has
    \kl{$\omega$-type} $(S, \vec{r})$.
    In particular, the following
    equality holds:
    \begin{equation*}
        f\left(\prod_{i = 1}^k 
            a_i^{ x_i \omega \ind{S}(i) + r_i}
        \right)
        =
        P_{(S, \vec{r})}(\seqof{x_i}{i \in S})
        \quad  \forall \vec{x} \in (\Nat_{\geq 1})^k
        \quad .
    \end{equation*}

>>>>

    Note that the same proof scheme can be used to conclude that
    \cref{f-npoly-combi:item} implies
    \cref{f-combinatorial:item}. For the converse implication, we are
    going to introduce $\omega_2$ associated to the fact that $f$ is
    \kl{$(k,\Nat)$-combinatorial}. Because polynomials \kl{represented} by
    \kl{$\Nat$-polyregular functions} and \kl{integer binomial polynomials} are
    both closed under multiplication of their input by a constant factor, we can assume that
    $\omega = \omega_2$ in the decomposition of $f$.
    Now, consider $(S, \vec{r}) \in \ModuloTypes[\omega]^k$. 
    Notice that for all vectors $\vec{x} \in (\Nat_{\geq 1})^k$, 
    the vector $(x_1 \omega \ind{S}(1) + r_1, \dots, x_k \omega \ind{S}(k) + r_k)$ has
    \kl{$\omega$-type} $(S, \vec{r})$.
    In particular, the following
    equality holds:
    \begin{equation*}
        f\left(\prod_{i = 1}^k 
            a_i^{ x_i \omega \ind{S}(i) + r_i}
        \right)
        =
        P_{(S, \vec{r})}(\seqof{x_i}{i \in S})
        \quad  \forall \vec{x} \in (\Nat_{\geq 1})^k
        \quad .
    \end{equation*}
